Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-04-06 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update for re-start project

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care \And Gerko Vink\\Methodology and
Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

@@

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}. Footnotes in the current
version show work in progress/under construction. The last section is
not part of the manuscript, but purely for reminders. See also all of
the TODOs that need to be worked out. We aim to submit at JSS, so there
is no word count limit (``There is no page limit, nor a limit on the
number of figures or tables''). {[}Just adding some text to get a better
guess of what the actura abstract will look like: Lorem ipsum dolor sit
amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut
labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud
exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
Duis aute irure dolor in reprehenderit in voluptate velit esse cillum
dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
proident, sunt in culpa qui officia deserunt mollit anim id est
laborum.{]}
}

@@

\Address{
    Hanne I. Oberman\\
    Methodology and Statistics\\
Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
        Thomas P. A. Debray\\
    Julius Center for Health Sciences and Primary Care\\
    University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\

@@

Many datasets include individuals that are clustered together, for
example in geographic regions, or even different studies. In the
simplest case, individuals (e.g., students) are nested within a single
cluster (e.g., school classes). More complex clustered structures may
occur when there are multiple hierarchical levels (e.g., students in
different schools or patients within hospitals within regions across
countries), or when the clustering is non-nested (e.g., electronic
health record data from diverse settings and populations within large
databases). With clustered data we generally assume that individuals
from the same cluster tend to be more similar than individuals from
other clusters. In statistical terms, this implies that observations
from the same cluster are not independent and may in fact be correlated.
If this correlation is left unaddressed, estimates of \emph{p} values,
confidence intervals even model parameters are prone to bias
\citep{loca01}. {[}TODO: make a link to imputation methods, which
require adequate handling and propagation of variance; we are not
recommending the adoption of multilevel models for data analysis here,
but rather for imputation. VMTdJ: ``I would only make that link after
introducing missing data, and missing data is not mentioned yet, so I
would do that in the next section''{]} Statistical methods for clustered
data typically adopt hierarchical models that explicitly describe the
grouping of observations. These models are also known as `multilevel
models', `hierarchical models', `mixed effect models', `random effect
models', and in the context of time-to-event data as `frailty models'.
Table \ref{tab:clus} provides an overview of some key concepts in
multilevel modeling.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels. A three-level \\
Fixed effect & Here we assume that the values of an independent variable
are fixed, i.e., \\
& the values observed in the study are representative of all values in
the \\
& in the dependent variable y e.g., blood pressure between treatments A
and B. \\
Random effect & The values of an independent variable are assumed to be
randomly drawn from \\
& admission we might select only certain hospitals that are
representative of \\
& the difference of y between individual hospitals, but rather the
variation of \\
ICC & The variability due to clustering is often measured by means of
the \\
& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
& of variance that can be attributed to the cluster-level, where a
high \\
& ICC would indicate that a lot of variability is due to the cluster \\
& structure. \\
Random effect & Multilevel models typically accommodate for variability
by including \\
& a separate group mean for each cluster. In addition to random \\
& intercepts, multilevel models can also include random coefficients \\
& and heterogeneous residual error variances across clusters {[}see
e.g. \\
& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
stratification.{]} \\
\bottomrule()
\end{longtable}

@@

As with any other dataset, clustered datasets may be impacted by
missingness in much the same way. Several strategies can be used to
handle missing data, including complete case analysis and imputation. We
focus on the latter approach and discuss statistical methods for
replacing the missing data with one or more plausible values. Imputation
seperates the missing data problem from the analysis and the completed
data can be analyzed as if it were completely observed. It is generally
recommended to impute the missing values more than once to preserve
uncertainty due to missingness and to allow for valid inferences (c.f.
Rubin 1976).

With incomplete clustered datasets we can distinguish between two types
of missing data: sporadic missingness and systematic missingness
\citep{resc13}. Sporadic missingness arises when variables are missing
for some but not all of the units in a cluster \citep{buur18, jola18}.
For example, it is possible that test results are missing for several
students in one or more classes. {[}TODO: Provide an example for one of
the case studies below.{]} When all observations are missing within one
or more clusters, data are said to be systematically missing. {[}TODO:
Refer to Figure 1 and put interpretation in the figure caption. VMTdJ:
``Mention in text that this is sporadic missingness?''{]}

@@

\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
\end{figure}
\end{CodeChunk}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& {[}TODO: add congeniality, but maybe in-text?{]} \\
\bottomrule()
\end{longtable}

{[}GV: DELETE THIS SECTION?{]}However, commonly used imputation methods
were not designed for use in clustered data and usually generate
observations that are independent whereas multilevel data are dependent.
For this reason, we discuss how the \proglang{R} package \pkg{mice} can
be used to impute multilevel data.

{[}TODO: clarify why clustering is relevant during imputation, and why
this exposes the need for specialized imputation methods and more
attention during their implementation (``thou shall not simply run
\texttt{mice()} on any incomplete dataset'').{]} {[}TODO: Add that the
more the random effects are of interest, the more you need multilevel
imputation models.{]} {[}TODO: Add an overview of all possible predictor
matrix values in manuscript or \texttt{ggmice} legend.{]}

@@

We provide practical guidelines and code snippets for different missing
data situations, including non-ignorable mechanisms. For reasons of
brevity, we focus on multilevel imputation by chained equations with
\pkg{mice} exclusively; other imputation methods and packages (e.g.,
\pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial.
{[}TODO: change this sentence because we do not assume familiarity:
``Assumed knowledge includes basic familiarity with multilevel
imputation \citep[see e.g.][ and \citet{grun18}]{audi18} and the
\pkg{lme4} notation for multilevel models (see Table
\ref{tab:mod})''{]}.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Formula \pkg{lme4}}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
\texttt{x1} predictor with random intercept \\
& varying among \texttt{g1} \\
\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
\texttt{x1} and \texttt{x2} only in fixed effect \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
of \texttt{x1} and \texttt{x2} only in fixed effect \\
& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
variance-covariance matrix estimated only with \\
& the variance terms of intercept, slope of \texttt{x1}, \\
& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
& variance-covariance matrix estimated separately, \\
& i.e, one for intercept and \texttt{x1} and another for \\
& intercept and \texttt{x2} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with correlated random intercept and \\
& random slope of \texttt{x} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with uncorrelated random intercept \\
& and random slope of \texttt{x1} \\
\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
& Random intercept varying among \texttt{g1} and among
\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
& {[}TODO: explain \pkg{lme4} notation here \\
\bottomrule()
\end{longtable}

@@

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
  with data that are MAR,][]{mice};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies with data that are MAR,][]{metamisc};
\item
  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions with data
  that are MNAR,][]{GJRM}.
\end{itemize}

@@

This tutorial is dedicated to readers who are unfamiliar with multiple
imputation. More experienced readers can skip the introduction (case
study 1) and directly head to practical applications of multilevel
imputation under MAR conditions (case study IMPACT) or under MNAR
conditions (case study HIV).

TODO: explicit statement about not going into workings of the methods.
Galimer 2l methods.

@@

\begin{itemize}
\tightlist
\item
  GV: CREATE A SINGLE ONLINE DEPARTURE POINT ON GITHUB?
\end{itemize}

{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
Set up the R environment and load the necessary packages:

@@

{[}TODO: explain case study{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis)
\end{CodeInput}
\begin{figure}

@@

{[}TODO: Add that it doesn't work with systematic missingness (only with
sporadic). There's some pros and cons, and it may not even differ much
if the number of clusters is low.{]}

@@

{[}TODO: check if there is systematic missingness in this dataset, if
not make Marshall Computerized Tomography classification (ct)
systematically missing.{]}

We illustrate how to impute incomplete multilevel data by means of a
case study: \texttt{impact} from the \pkg{metamisc} package
\citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
across \(N = 15\) clusters,][]{metamisc}. {[}TODO: add more info about
the complete data.{]} The \texttt{impact} data set contains traumatic
brain injury data on \(n = 11022\) patients clustered in \(N = 15\)
studies with the following 11 variables:

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification, {[}TODO:
  make this one var? also shows that you don't always need random
  effects everywhere?{]}
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

{[}TODO: show how much variance there is after different methods{]}

{[}TODO: add ICC before/after imputation and interpret: This tells us
that the multilevel structure of the data should probably be taken into
account. If we don't, we'll may end up with incorrect imputations,
biasing the effect of the clusters towards zero.{]}

{[}TODO: add descriptive statistics of the complete and incomplete
data.{]}

@@

To explore the missingness, it is wise to look at the missing data
pattern:

@@

This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
and \texttt{pupil}.{]}

To develop the best imputation model, we need to investigate the
relations between the observed values of the incomplete variables and
the observed values of other variables, and the relation between the
missingness indicators of the incomplete variables and the observed
values of the other variables. To see whether the missingness depends on
the observed values of other variables, we\ldots{} {[}TODO: fill in that
we can test this statistically or use visual inspection (e.g.~a
histogram faceted by the missingness indicator).{]}

@@

\hypertarget{complete-case-analysis-todo-remove-this}{%
\subsection{Complete case analysis {[}TODO: remove
this?{]}}\label{complete-case-analysis-todo-remove-this}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   3.00e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.56e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

As we can see\ldots{} {[}TODO: fill in: This means that a higher ct
(Marshall Computerized Tomography classification) is associated with a
lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI
\ldots{} to \ldots, when controlling for \ldots{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(dat) # methods vector
R> pred <- quickpred(dat)   # predictor matrix
R> plot_pred(pred)
\end{CodeInput}

@@

\begin{CodeInput}
R> pred[pred == 1] <- 2
R> pred["mort", ] <- 2
R> pred[, "mort"] <- 2
R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
R> pred[, "name"] <- -2
R> diag(pred) <- 0
R> plot_pred(pred)
\end{CodeInput}

@@

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10505.026 10570.795 -5243.513 10487.026     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2908  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32339                 -0.42359                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67142                 -1.05776                 -1.51038  
                  pupil                       ct  
                0.49756                  0.42474  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

>>>>

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

@@

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}.
}

@@

\Address{
    Hanne I. Oberman\\
    Methodology and Statistics\\
Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
          }

@@

Many datasets include individuals that are clustered together, for
example in geographic regions, or even different studies. In the
simplest case, individuals (e.g., students) are nested within a single
cluster (e.g., school classes). More complex clustered structures may
occur when there are multiple hierarchical levels (e.g., students in
different schools or patients within hospitals within regions across
countries), or when the clustering is non-nested (e.g., electronic
health record data from diverse settings and populations within large
databases). With clustered data we generally assume that individuals
from the same cluster tend to be more similar than individuals from
other clusters. In statistical terms, this implies that observations
from the same cluster are not independent and may in fact be correlated.
If this correlation is left unaddressed, estimates of \emph{p} values,
confidence intervals even model parameters are prone to bias
\citep{loca01}. Statistical methods for clustered data typically adopt
hierarchical models that explicitly describe the grouping of
observations. These models are also known as `multilevel models',
`hierarchical models', `mixed effect models', `random effect models',
and in the context of time-to-event data as `frailty models'. Table
\ref{tab:clus} provides an overview of some key concepts in multilevel
modeling.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels. A three-level \\
Fixed effect & Here we assume that the values of an independent variable
are fixed, i.e., \\
& the values observed in the study are representative of all values in
the \\
& in the dependent variable y e.g., blood pressure between treatments A
and B. \\
Random effect & The values of an independent variable are assumed to be
randomly drawn from \\
& admission we might select only certain hospitals that are
representative of \\
& the difference of y between individual hospitals, but rather the
variation of \\
ICC & The variability due to clustering is often measured by means of
the \\
& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
& of variance that can be attributed to the cluster-level, where a
high \\
& ICC would indicate that a lot of variability is due to the cluster \\
& structure. \\
Random effect & Multilevel models typically accommodate for variability
by including \\
& a separate group mean for each cluster. In addition to random \\
& intercepts, multilevel models can also include random coefficients \\
& and heterogeneous residual error variances across clusters {[}see
e.g. \\
& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
stratified intercept as concept..{]} \\
\bottomrule()
\end{longtable}

@@

As with any other dataset, clustered datasets may be impacted by
missingness in much the same way. Several strategies can be used to
handle missing data, including complete case analysis and imputation. We
focus on the latter approach and discuss statistical methods for
replacing the missing data with one or more plausible values. Imputation
separates the missing data problem from the analysis and the completed
data can be analyzed as if it were completely observed. It is generally
recommended to impute the missing values more than once to preserve
uncertainty due to missingness and to allow for valid inferences (c.f.
Rubin 1976).

With incomplete clustered datasets we can distinguish between two types
of missing data: sporadic missingness and systematic missingness
\citep{resc13}. Sporadic missingness arises when variables are missing
for some but not all of the units in a cluster \citep{buur18, jola18}.
For example, it is possible that test results are missing for several
students in one or more classes. When all observations are missing
within one or more clusters, data are said to be systematically missing.

@@

\caption[Sporadic missingness in multilevel data]{Sporadic missingness in multilevel data}\label{fig:patterns}
\end{figure}
\end{CodeChunk}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& \\
\bottomrule()
\end{longtable}

@@

We provide practical guidelines and code snippets for different missing
data situations, including non-ignorable mechanisms. For reasons of
brevity, we focus on multilevel imputation by chained equations with
\pkg{mice} exclusively; other imputation methods and packages \citep[see
e.g.][ and \citet{grun18}]{audi18} are outside the scope of this
tutorial. Assumed knowledge includes basic familiarity with the
\pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Formula \pkg{lme4}}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
\texttt{x1} predictor with random intercept \\
& varying among \texttt{g1} \\
\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
\texttt{x1} and \texttt{x2} only in fixed effect \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
of \texttt{x1} and \texttt{x2} only in fixed effect \\
& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
variance-covariance matrix estimated only with \\
& the variance terms of intercept, slope of \texttt{x1}, \\
& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
& variance-covariance matrix estimated separately, \\
& i.e, one for intercept and \texttt{x1} and another for \\
& intercept and \texttt{x2} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with correlated random intercept and \\
& random slope of \texttt{x} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with uncorrelated random intercept \\
& and random slope of \texttt{x1} \\
\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
& Random intercept varying among \texttt{g1} and among
\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
& \\
\bottomrule()
\end{longtable}

@@

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
  with data that are MAR,][]{mice};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies with data that are MAR,][]{metamisc};
\item
  \texttt{obesity} from the \pkg{miceheckman} package {[}simulated data
  on obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
  that are MNAR{]}.
\end{itemize}

@@

This tutorial is dedicated to readers who are unfamiliar with multiple
imputation. More experienced readers can skip the introduction (case
study 1) and directly head to practical applications of multilevel
imputation under MAR conditions (case study 2) or under MNAR conditions
(case study 3).

@@

Install non-CRAN packages if necessary:

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("amices/ggmice")
R> devtools::install_github("hanneoberman/miceheckman")
\end{CodeInput}
\end{CodeChunk}

Set up the R environment and load the necessary packages:

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis, square = TRUE)
\end{CodeInput}
\begin{figure}

@@

@@

We illustrate how to impute incomplete multilevel data by means of a
case study: \texttt{impact} from the \pkg{metamisc} package
\citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
across \(N = 15\) clusters,][]{metamisc}. The \texttt{impact} data set
contains traumatic brain injury data on \(n = 11022\) patients clustered
in \(N = 15\) studies with the following 11 variables:

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

@@

To explore the missingness, it is wise to look at the missing data
pattern. The ten most frequent missingness patterns are shown:

@@

This shows that we need to impute \texttt{ct} and \texttt{pupil}.

To develop the best imputation model, we need to investigate the
relations between the observed values of the incomplete variables and
the observed values of other variables, and the relation between the
missingness indicators of the incomplete variables and the observed
values of the other variables. To see whether the missingness depends on
the observed values of other variables, we can test this statistically
or use visual inspection (e.g.~a histogram faceted by the missingness
indicator).

@@

\hypertarget{complete-case-analysis}{%
\subsection{Complete case analysis}\label{complete-case-analysis}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

As we can see, a higher \texttt{ct} (Marshall Computerized Tomography
classification) is associated with a lower odds of 6-month mortality,
given by the odds ratio exp(0.42), CI \ldots{} to \ldots, when
controlling for\ldots{}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(dat) # methods vector
R> pred <- quickpred(dat)   # predictor matrix
R> plot_pred(pred, rotate = TRUE)
\end{CodeInput}

@@

\begin{CodeInput}
R> pred[pred == 1] <- 2
R> pred["mort", ] <- 2
R> pred[, "mort"] <- 2
R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
R> pred[, "name"] <- -2
R> diag(pred) <- 0
R> plot_pred(pred, rotate = TRUE)
\end{CodeInput}

@@

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10505.026 10570.795 -5243.513 10487.026     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2908  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32327                 -0.42365                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67143                 -1.05776                 -1.51039  
                  pupil                       ct  
                0.49756                  0.42474  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.511 10585.280 -5250.755 10501.511     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2961  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33576                 -0.40872                  0.03039  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66477                 -1.05453                 -1.51860  
                  pupil                       ct  
                0.45928                  0.44419  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
